Time for a rethink: time sub-sampling methods in disparity-through-time analyses

### 2017-12-14 manuscript/time-slice.tex
edits on the manuscript up until time-slicing.

Disparity-through-time analyes are common in palaeontology \citep{Foote01071994}.
They reveal how the morphological diversity of clades has changed through time, and allow us to make inferences about the breadth of ecological niches extinct taxa occupied \citep{foote1997evolution}.
Results from disparity-through-time studies also provide insights into the ecological impacts of mass extinctions, competitive replacements, and the drivers of morphological evolution \citep{Brusatte12092008,Foote29111996,friedmanexplosive2010}.
Unfortunately, the way we perform these analyses may have profound effects on our conclusions.

Disparity-through-time analyes have two main analysis components: calculating disparity, and creating time sub-samples of the data. 
Here we focus on the latter.
The nature of disparity (it is a diversity metric), means it cannot be calculated using just one individual, so some way of subsetting taxa is required.
Changes in disparity-through-time are generally investigated by calculating the disparity of taxa present during specific time intervals or time bins \citep[e.g][]{cisneros2010,prentice2011,Hughes20082013,hopkinsdecoupling2013,bentonmodels2014,bensonfaunal2014}.
These time bins are usually defined based on stratigraphy \citep[e.g.][]{cisneros2010,prentice2011,Hughes20082013,bentonmodels2014} but can also be arbitrarily chosen time bins of equal duration \citep{Butler2012,hopkinsdecoupling2013,bensonfaunal2014}.
However, this approach has several limitations.

First, time bins defined by stratigraphy are not of equal size, biasing higher disparity towards longer stratigraphic periods. 
This can be dealt with using rarefaction, i.e. repeating the analysis using the smallest number of taxa present in any bin as the number of sampled taxa in each bin. 
This can, however, lead to very large confidence intervals when there are stratigraphic periods with very few species.
Other studies split large time bins so they are of roughly equal size, but this is often an \textit{ad hoc} procedure that can introduce more bias depending on where bins are split.
Second, all time binning approaches (whether bins are equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring \textit{between} intervals, rather than also allowing for gradual changes \textit{within} intervals \citep{Hunt21042015}.
Finally, time bin analyses are often limited by the number of taxa in each bin.
If there are insufficient taxa in a time bin, disparity cannot be calculated, so further analyses, e.g. correlations of disparity with hypothesised drivers of morphological evolution, are not possible.

To address these issues, we propose a ``time-slicing'' approach that takes advantage of the wealth of paleontological datasets which now have associated phylogenies. 
Time-slicing uses a phylogenetic tree and considers subsets of taxa at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
This results in even-sampling across time and permits us to define the underlying model of character evolution (punctuated or gradual).  
Time-slicing also includes any element present in the phylogeny (branches, nodes and tips) at the time slice in question as part of the disparity calculation.
This allows us to measure disparity at time points where there are no terminal taxa, and increases the sample size at each time point, making downstream analyses of the drivers of disparity much more feasible.

@@

\section{Materials and Methods}
\subsection{Overview}
\label{overview-section}
To test the different time sub-sampling methods, we followed the protocol below (Figure \ref{datasets}). 
All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (https://github.com/nhcooper123/time-slice).

%https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/details-of-specific-functions.html#time-slicing$

@@

\subsection{Example datasets}
\label{datasets}
To test the different time binning/slicing methods we selected four datasets: a mammal dataset from \cite{beckancient2014}, two theropod datasets from \cite{brusatte2014gradual} and \cite{bapst2016topology}, and a crinoid dataset from \cite{wright2017bayesian}.
Table \ref{table:datasets} and the Supporting Information Appendix S1 provide more details. 
Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
These datasets are freely available with their accompanying papers (Table \ref{table:datasets}), but for reproducibility purposes we also provide the data we used on GitHub (https://github.com/nhcooper123/time-slice/data).

@@

\paragraph{Estimating ancestral character states.}
\label{ace}
For each dataset we used the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny, using the \texttt{rerootingMethod} function from the \texttt{R} package \texttt{phytools} version 0.6-44 \citep{phytools,R}.
Where there were missing character data for a taxon we followed the method of \cite{Claddis} and treated missing data as any possible observed state for each character.
For example, if a character had two observed states ($0$ and $1$) across all taxa, we attributed the multi-state ``$0$\&$1$" value to the taxon with missing data, representing an equal probability of being either $0$ or $1$.
This allows the ancestral node of a taxon with missing data to be estimated with no assumptions other than that the taxon has one of the observed character states.
To prevent poor ancestral state reconstructions from biasing our results, especially when a lot of error is associated with the reconstruction, we only included ancestral state reconstructions with a scaled Likelihood $\geq$ $0.95$.
Ancestral state reconstructions with scaled Likelihoods below this threshold were recoded as missing data (``?'').

\paragraph{Building cladisto-spaces.} 
To explore disparity-through-time in our datasets we used a cladisto-space approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
This approach is similar to constructing a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010}, except a cladisto-space is an approximation of the morphospace based on cladistic data (i.e. the discrete morphological characters used to build a phylogenetic tree).
Mathematically, a cladisto-space is an $n$ dimensional object that summarises the cladistic distances between the taxa present in a cladistic matrix.
Although empirically inter-taxon distances are the same in a morphospace or a cladisto-space \citep{foth2012different,hetherington2015cladistic}, we prefer the term cladisto-space to make it clear that this space is estimated using cladistic data and not morphometric data.

@@

\paragraph{Ordination.}
After constructing our distance matrices we transformed them using classical multidimensional scaling \citep[MDS;][]{torgerson1965multidimensional,GOWER01121966,cailliez1983analytical}.
This method (also referred to as PCO; e.g. \citealt{Brusatte2015}; or PCoA; e.g. \citealt{paradisape:2004}) is an eigen decomposition of the distance matrix.
Because we used Gower distances instead of raw Euclidean distances, negative eigenvalues can be calculated.
To avoid this problem, we first transformed the distance matrices by applying the Cailliez correction \citep{cailliez1983analytical} which adds a constant $c^*$ to the values in a distance matrix (apart from the diagonal) so that all the Gower distances become Euclidean ($d_{Gower}+c^*=d_{Euclidean}$; \citealt{cailliez1983analytical}). 
We were then able to extract $n$ eigenvectors for each matrix (representing the $n$ dimensions of the cladisto-space) where $n$ is equal to $k-2$, i.e. the number of taxa in the matrix ($k$) minus the last two eigenvectors that are always null after applying the Cailliez correction.
Contrary to previous studies \citep[e.g][]{brusatte50,cisneros2010,prentice2011,anderson2012using,Hughes20082013,bentonmodels2014}, we use all $n$ dimensions of our cladisto-spaces and not a sub-sample representing the majority of the variance in the distance matrix (e.g. selecting only $m$ dimensions that represent up to 90\% of the variance in the distance matrix; \citealt{Brusatte12092008,toljagictriassic-jurassic2013}).

@@

@@

@@

\section{Acknowledgments}
	NC thanks Mark Sutton and Phillip Mannion for the invitation to contribute to the `Evolutionary Modelling' symposium at The Palaeontological Association Annual Meeting 2017.
  TG acknowledges support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.
  We thank Dave Bapst, Steve Brusatte, Graeme Lloyd, April Wright and Davey Wright for assistance in gathering data for the analyses.

>>>>

Disparity-through-time analyes are common in palaeontology since the 90s \citep{gould1991disparity,briggs1992morphological,Wills1994,Foote01071994}.
They reveal how the morphological diversity of clades has changed through time, and allow us to make inferences about the breadth of ecological niches extinct taxa occupied \citep{foote1997evolution}.
Results from disparity-through-time studies also provide insights into the ecological impacts of mass extinctions, competitive replacements, and the drivers of morphological evolution \citep{Brusatte12092008,Foote29111996,friedmanexplosive2010}.
Unfortunately, the way we perform these analyses may have profound effects on our conclusions.

Disparity-through-time analyes have two main analysis components: calculating disparity, and creating time sub-subsets of the data. 
Here we focus on the latter.
The nature of disparity (i.e. a diversity metric), means it cannot be calculated using a single individual, so some way of sampling many taxa in each sub-subset is required.
Changes in disparity-through-time are generally investigated by calculating the disparity of taxa present during specific time intervals or time bins \citep[e.g][]{cisneros2010,prentice2011,Hughes20082013,hopkinsdecoupling2013,bentonmodels2014,bensonfaunal2014}.
These time bins are usually defined based on stratigraphy \citep[e.g.][]{cisneros2010,prentice2011,Hughes20082013,bentonmodels2014} but can also be arbitrarily chosen time bins of equal duration \citep{Butler2012,hopkinsdecoupling2013,bensonfaunal2014}.
However, this approach has several limitations.

First, time bins defined by stratigraphy are not of equal size, biasing higher disparity towards longer stratigraphic periods. 
This can be dealt with using rarefaction methods, i.e. repeating the analysis while re-sampling the taxa to have the same amount of taxa in each bin (e.g. using bootstrapping with limited re-sampling methods).
This can, however, lead to very large confidence intervals when there are stratigraphic periods with very few species.
Other studies split large time bins so they are of roughly equal size, but this is often an \textit{ad hoc} procedure that can introduce more bias depending on where bins are split.
Second, all time binning approaches (whether bins are equally sized or not) assume that characters of taxa evolve following a punctuated equilibrium model, because disparity is only estimated once for each interval resulting in all changes in disparity occurring \textit{between} intervals, rather than also allowing for gradual changes \textit{within} intervals \citep[shown as being quiet common in the fossil record;][]{Hunt21042015}.
Finally, time bin analyses are often limited by the number of taxa in each bin.
If there are insufficient taxa in a time bin, disparity cannot be calculated, so further analyses, e.g. correlations of disparity with hypothesised drivers of morphological evolution, are not possible.
%TG: Not sure where in this paragraph but it'll be good to add this point:
%Also, in the case of looking at changes in disparity due to events specific in time (e.g. a mass extinction), time bins method leave us with a low temporal resolution: stratigraphic bins are limited to global events that defined the bins boundaries and time bins are limited by the duration of each bin (e.g. if time bins are every 20 Mya but an event occurs at time $t$ and $t+5$ Mya, it might not be possible to capture its effects).

To address these issues, we propose a ``time-slicing'' approach that takes advantage of the wealth of paleontological datasets which now have associated phylogenies. 
Time-slicing uses a phylogenetic tree and considers subsets of taxa at specific equidistant points in time, as opposed to considering subsets of taxa between two points in time.
This results in even-sampling across time and permits us to define the underlying model of character evolution (punctuated or gradual).  
Time-slicing also includes any element present in the phylogeny (branches, nodes and tips) at the time slice in question as part of the disparity calculation.
This allows us to measure disparity at time points where there are no sampled terminal taxa, and increases the sample size at each time point, making downstream analyses of the drivers of disparity much more feasible.

@@

\section{Materials and Methods}
\subsection{Overview}
\label{overview-section}
To test the different time sub-sampling methods, we followed the protocol below (Figure \ref{datasets}). 
All the code needed to reproduce these analyses (along with detailed instructions) is provided on GitHub (\url{https://github.com/nhcooper123/time-slice}).
%https://rawgit.com/TGuillerme/dispRity/master/inst/gitbook/_book/details-of-specific-functions.html#time-slicing$

@@

\subsection{Example datasets}
\label{datasets}
To test the different time binning/slicing methods we selected four datasets: a mammal dataset from \cite{beckancient2014}, two theropod datasets from \cite{brusatte2014gradual} and \cite{bapst2016topology}, and a crinoid dataset from \cite{wright2017bayesian}.
Table \ref{table:datasets} and the Supporting Information Appendix S1 provide more details. 
Each dataset consists of first and last occurrence dates for all taxa, a matrix of morphological characters in NEXUS format, and a time-scaled phylogeny. 
These datasets are freely available with their accompanying papers (Table \ref{table:datasets}), but for reproducibility purposes we also provide the data we used on GitHub (\url{https://github.com/nhcooper123/time-slice/data}).

@@

\paragraph{Estimating ancestral character states.}
\label{ace}
For each dataset we estimated the ancestral character states at each node using the \texttt{AncStatesEstMatrix} function from the \texttt{Claddis} \texttt{R} package \citep{Claddis,R}.
This function uses the re-rooting method \citep{Yang01121995,Garland2000} to get Maximum Likelihood estimates of the ancestral states for each character at every node in the phylogeny \citep[based on the \texttt{rerootingMethod} function in \texttt{phytools};][]{phytools}.
% Inapplicable and missing character for any taxon were treated as ambiguous characters (i.e. any possible observed state for the character).
% For example, if a character had two observed states ($0$ and $1$) across all taxa, we attributed the multi-state ``$0$\&$1$" value to the taxon with missing data, representing an equal probability of being either $0$ or $1$.
% This allows the ancestral node of a taxon with missing data to be estimated with no assumptions other than that the taxon has one of the observed character states.
%TG: These three sentences can be removed: it's pretty standard procedure. I'm not sure if we should leave that here. It sounded fancy during PhD time but now it's actually a bit lame to write it like that. Plus there is the inapplicable problem that Claddis do not deal with...
To prevent poor ancestral state reconstructions from biasing our results, especially when a lot of error is associated with the reconstruction, we only included ancestral state reconstructions with a scaled Likelihood $\geq$ $0.95$.
Ancestral state reconstructions with scaled Likelihoods below this threshold were recoded as missing data (``?'').
This allowed our results to be less dependent on the quality (or the absence thereof) of the ancestral state reconstruction methods, especially in parts of the datasets where data was sparse.

\paragraph{Building cladisto-spaces.} 
To explore disparity-through-time in our datasets we used a cladisto-space approach \citep[e.g.][]{Foote01071994,Foote29111996,Wesley-Hunt2005,Brusatte12092008,friedmanexplosive2010,toljagictriassic-jurassic2013,Hughes20082013}.
This approach is similar to constructing a morphospace based on continuous morphological data \citep[e.g.][]{friedmanexplosive2010}, except a cladisto-space is an approximation of the morphospace based on discrete morphological data (i.e. the discrete morphological characters used to build a phylogenetic tree, often referred to as ``cladistic'' data).
%TG: about that, I'm actually less and less convinced we should call it caldisto-space: first, semantically, cladistic data means data used to make clades (phylos) so it could be nucleotides as well. Second, during the PhD, we used cladisto-space to distinguish it from the morpho-space but at that time I was naively thinking the the morpho-space is the morphometric-space (which it often is!) but, again, semantically, the morpho-space can also be simply the morphology-space... However, I like the sound of cladisto-space and most people will understand it the way we do. I'm just thinking about pedantic-palaeo-reviewers...
Mathematically, a cladisto-space is an $n$ dimensional object that summarises the cladistic distances between the taxa present in a cladistic matrix.
Although empirically inter-taxon distances are the same in a morphospace or a cladisto-space \citep{foth2012different,hetherington2015cladistic}, we prefer the term cladisto-space to make it clear that this space is estimated using cladistic data and not morphometric data.

@@

\paragraph{Ordination.}
After constructing our distance matrices we transformed them using classical multidimensional scaling \citep[MDS;][]{torgerson1965multidimensional,GOWER01121966,cailliez1983analytical}.
This method (also referred to as PCO; e.g. \citealt{Brusatte2015}; or PCoA; e.g. \citealt{paradisape:2004}; but distinguished in \citealt{legendre2012numerical}) is an eigen decomposition of the distance matrix.
Because we used Gower distances instead of raw Euclidean distances, negative eigenvalues can be calculated.
To avoid this problem, we first transformed the distance matrices by applying the Cailliez correction \citep{cailliez1983analytical} which adds a constant $c^*$ to the values in a distance matrix (apart from the diagonal) so that all the Gower distances become Euclidean ($d_{Gower}+c^*=d_{Euclidean}$; \citealt{cailliez1983analytical}). 
We were then able to extract $n$ eigenvectors for each matrix (representing the $n$ dimensions of the cladisto-space) where $n$ is equal to $k-2$, i.e. the number of taxa in the matrix ($k$) minus the last two eigenvectors that are always null after applying the Cailliez correction.
Contrary to previous studies \citep[e.g][]{brusatte50,cisneros2010,prentice2011,anderson2012using,Hughes20082013,bentonmodels2014}, we use all $n$ dimensions of our cladisto-spaces and not a sub-sample representing the majority of the variance in the distance matrix (e.g. selecting only $m$ dimensions that represent up to 90\% of the variance in the distance matrix; \citealt{Brusatte12092008,toljagictriassic-jurassic2013}).

@@

The disparity-through-time analyses were performed using the \texttt{dispRity} \texttt{R} package \citep{dispRity.v02}. %TG: With some hope, the MEE paper might be out at the same time as this one!

@@

%TG: Is it worth saying that Halliday and Goswami came up with a similar idea \citep{halliday2016eutherian} but that we came up with it first \cite{Guillerme2015} and we're the first to implement it \cite{dispRity.v02}?

@@

\section{Acknowledgments}
	NC thanks Mark Sutton and Phillip Mannion for the invitation to contribute to the `Evolutionary Modelling' symposium at The Palaeontological Association Annual Meeting 2017.
  TG acknowledges support from the Australian Discovery Project Grant number DP170103227 awarded to Vera Weisbecker.
  We thank Dave Bapst, Steve Brusatte, Graeme Lloyd, April Wright and Davey Wright for assistance in gathering data for the analyses.
  %TG: and actually: Dave Bapst, Graeme Lloyd and Davey Wrigth for discussions about the method?
